---
title: "Concepts on cell types and signatures"
author:
- name: Kevin Rue-Albrecht
  affiliation:
  - &id1 Kennedy Institute of Rheumatology, University of Oxford, Headington, Oxford OX3 7FY, UK.
  email: kevin.rue-albrecht@kennedy.ox.ac.uk
- name: Second Author
  affiliation: Second Author's Affiliation
  email: corresponding@author.com
package: Hancock
output:
    BiocStyle::html_document
abstract: |
    A discussion of concepts associated with cell types and transcriptional signatures.
vignette: |
    %\VignetteIndexEntry{Concepts on cell types and signatures}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

# Definition of cell identity

Discuss:

- cell type is a continuum (e.g. differentiation, pseudotime)
- in the same vein, all cells in an organism basically originate from a common progenitor
- cell states may be defined similarly to cell types (e.g. markers of activated/resting cells)

# Definition of absolute and relative markers and signatures

## Definition

**Absolute markers** may be defined as molecules known to be present or absent in a given population of cells, _irrespective of their expression in other cells of the same sample_.
For instance, T helper lymphocytes can be defined by the presence of surface proteins Cd3 and Cd4, while T cytotoxic lymphocytes are defined by the presence of surface proteins Cd3 and Cd8.
To assess whether a cell is likely a T helper lymphocyte, one does not need to know the markers of T cytotoxic lymphocytes, nor that the two cell types have the Cd3 marker in common.

**Relative markers** may be defined by _differential_ expression against other cells in the same sample.
For instance, in a biological sample including both T helper and T cytotoxic lymphocytes, differential expression between the two lypmphocyte subsets would highlight Cd4 as a (relative) marker of T helper lymphocytes and Cd8 as a marker of T cytotoxic lymphocytes.

Note that sets of **absolute markers** may also be subsetted to markers _specific_ of each cell population, either to increase stringency (due to their specificity) or sensitivity (due to their smaller number).
For instance, Cd3 being a marker of both T helper and T cytotoxic lymphocytes, one may wish to exclude it from both signatures, in a manner similar to **relative markers**.
Such signatures may be called **relative absolute signatures**, as they are composed of **absolute markers** compared to one another in order to identify specific subsets of markers.

## Applications

Absolute markers present the advantage of allowing the immediate characterization of any cell or cluster, without the need of a reference cell of cluster.
In this case, each cell or cluster may be screened for the presence of absolute markers, and assigned an identity independently of all other cells in the sample.

Relative markers can be advtantageous when the general cell type composition of a given sample is known in advance, and the problem is only to assign a predefined set of identities to a similar number of cell clusters in a new sample.
In this case, differential expression analysis may be performed between clusters of cells in the new sample, and markers of each cluster may be compared to a similar reference sample to assign identities defined in the reference sample to each of the cells or clusters in the new sample.
However, cells within a given sample are generally the result of sorting (e.g. FACS) and enriching a population of interest on a set of (protein) markers.
In those cases, the markers used for sorting the cells generally appear as highly expressed in all cells, making it difficult or impossible to identify as relative markers.

# Representation of signatures

At its simplest a signature could list _positive_ markers that are known to be expressed in each cell type.
For instance, the PBMC signatures used in the Seurat [PBMC 3k tutorial](https://satijalab.org/seurat/pbmc3k_tutorial.html) may be represented as follows:

```{r}
pbmc3k_markers <- list(
    "CD4 T cells" = c("IL7R"),
    "CD14+ Monocytes" = c("CD14", "LYZ"),
    "B cells" = c("MS4A1"),
    "CD8 T cells" = c("CD8A"),
    "FCGR3A+ Monocytes" = c("FCGR3A", "MS4A7"),
    "NK cells" = c("GNLY", "NKG7"),
    "Dendritic Cells" = c("FCER1A", "CST3"),
    "Megakaryocytes" = c("PPBP")
)
```


# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
