---
title: "Predicting gender"
author: "Kevin Rue-Albrecht"
date: "07/12/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the data.

```{r}
library(ExperimentHub)
ehub <- ExperimentHub()
EH1 <- ehub[["EH1"]]
library(SummarizedExperiment)
EH1 <- as(EH1, "SummarizedExperiment")
EH1
```

Tabulate numbers of known male and female subjects.

```{r}
table(EH1$gender)
```

Normalize in CPM.

```{r}
libSize <- colSums(assay(EH1, "exprs"))
assay(EH1, "CPM") <- t(t(assay(EH1, "exprs")) / libSize) * 1E6
```

Define a signature including all genes on the Y chromosome.

```{r}
library(EnsDb.Hsapiens.v86)
columns(EnsDb.Hsapiens.v86)
Ygenes <- unique(select(EnsDb.Hsapiens.v86, keys = "Y", "SYMBOL", "SEQNAME"))
Ygenes <- Ygenes$SYMBOL
table(Ygenes %in% rownames(EH1))
```

Let us use only the Y chromosome genes that are present in the data set.

```{r}
Ygenes <- intersect(Ygenes, rownames(EH1))
```

```{r}
library(GSEABase)
pbmc3k_markers_gsc <- GeneSetCollection(list(
    GeneColorSet(
        setName="Y chromosome expression in male", Ygenes, phenotype=c("Male"),
        geneColor=factor(rep(TRUE, length(Ygenes))),
        phenotypeColor=factor(rep(TRUE, length(Ygenes)))
    ),
    GeneColorSet(
        setName="Y chromosome expression in female", Ygenes, phenotype=c("Female"),
        geneColor=factor(rep(FALSE, length(Ygenes))),
        phenotypeColor=factor(rep(FALSE, length(Ygenes)))
    )
))
pbmc3k_markers_gsc
```


# Deprecate

Aggregate counts across all genes on the Y chromosome

```{r}
yCPM <- colSums(assay(EH1, "CPM")[Ygenes, ])
```

```{r}
# Cluster on the pairwise distance between total Y-chromosome CPM.
hc <- hclust(dist(yCPM))
plot(hc, labels=FALSE)
```

K-mean clustering with k=2 (expected number of genders)

```{r}
km <- kmeans(matrix(yCPM, ncol = 1), 2)
```

Test if the total Y-CPM of the two groups is significantly different

```{r}
testDataFrame <- data.frame(yCPM=yCPM, kmean=km$cluster)
ttOut <- t.test(yCPM ~ kmean, testDataFrame)
if (ttOut$p.value < 1E-3) {
    # there are both genders
    clusterMeans <- with(testDataFrame, tapply(yCPM, kmean, "mean"))
    clusterFemale <- which.min(clusterMeans)
    clusterMale <- which.max(clusterMeans)
    # Annotate known gender
    testDataFrame[colnames(EH1), "gender_annotated"] <- EH1$gender
    # Annotate predicted gender
    testDataFrame$gender_predicted <- factor(c("MALE", "FEMALE")[c(clusterMale, clusterFemale)][testDataFrame$kmean])
    # return the group with the highest km$centers as male
    # return the group with the highest km$centers as female
} else {
    # there is only one gender
    # we can only tell which gender if we're given an idea of what either "low" or "high" means
}
# Tabulate the predictions against the annotations
with(testDataFrame, table(gender_annotated, gender_predicted))
```

A large proportion of male subjects are classified as female.
How does the distribution of summed CPM look between genders?

```{r}
yCpmDataFrame <- data.frame(
    yCPM=yCPM,
    gender=colData(EH1[, names(yCPM)])[, "gender"]
)
require(ggplot2)
require(cowplot)
ggplot(yCpmDataFrame, aes(gender, yCPM)) +
    geom_point()
```

